A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-12-17 article/03_Results.Rmd
added some literature for the comparison

General observations: **TODO**
Literature! **TODO**

`r a("Southeastern Central Europe")`^[Amount of graves in Radon-B for this region.]:
The artificial study region *Southeastern Central Europe* comprises the eastern part of Austria (Upper Austria, Lower Austria, Styria, Burgenland), the northwest of Hungary, Western Slovakia, Moravia and large parts of Bohemia (Figure \ref{map} B). The density of ^14^C data on graves in Radon-B is low -- individual dates are available from the various regions, but large areas are without evidence. Diachronically the amount of data is consistently low and reaches a minimum in the Middle Bronze Age. While for the *burial type* variable a constant flow of information is given, the number of graves without information on *burial construction* increases in the Late Bronze Age to such an extent that no meaningful overview can be derived. The comparison with archaeological observations is complicated by the fact that the region intersects diverse geographical and cultural areas. In the southeast of Central Europe, cremation and bodily burial already existed side by side in the Early Bronze Age, and the local tradition remained stable in this respect also due to the expansion of the burial mound tradition in the Middle Bronze Age -- with a slight overall increase in cremation graves. The Urnfield period brought the transition to the absolute dominance of cremation burial. The development of the *burial type* variable in the Radon-B proxy shows this development correctly. The information from ^14^C data on *burial construction* correctly traces the relevance of flat burials in the Early Bronze Age, but the latter are then overemphasized in the Middle Bronze Age compared to the construction of burial mounds. In the Late Bronze Age, burials on flat burial grounds were frequent, although burial mounds of the Tumulus Culture were also further documented. This impact cannot be correctly deduced from Radon-B -- certainly due to the small amount of data.

`r a("Poland")`:
The region *Poland* completely covers the centre and western part of today's Poland. However, the ^14^C data from this region are almost exclusively from the voivodships of Kajuwien-Pommern, Lodsch and Lower Silesia. The amount of data is small overall, with significantly more data from the Early Bronze Age up to about 1700calBC. Almost no information is available from the Middle and Late Bronze Ages. This applies to both the *burial type* and the *burial construction* variables, the former being slightly better documented. The relative development of *burial type* approximates the *real* conditions -- i.e. according to the classical, archaeological narrative: inhumations were actually significantly more frequent in western Poland until the Late Bronze Age, while cremations did not occur until the Urnfield Period. Also before, cremation graves had been created sometimes, but in general the tendency recorded by Radon-B is correct. In contrast, *burial construction* is misrepresented: burial mounds also played an important role in the Middle Bronze Age.

`r a("Southern Germany")`:
The artificial district region *Southern Germany* covers Bavaria completely, but also includes most of Baden-Württemberg and Thuringia as well as extensive areas in Moravia and Western Austria. No ^14^C data are available from the latter regions. The majority of the available data comes from the Upper Palatinate, Lower Bavaria and Swabia as well as (in Württemberg) the administrative district of Tübingen. For the Early Bronze Age a remarkable amount of data from Southern Germany are preserved in Radon-B, but this amount decreases in the course of the Middle Bronze Age and leaves the Late Bronze Age almost without dated graves. The *burial type* variable is quite well understood -- although disproportionately much more of the early body graves were recorded than urnfield burials. This wealth of information contrasts sharply with the situation regarding the *burial construction* variable, for which almost no information from this area is available throughout the Bronze Age. If one considers the relative development calculated from Radon-B and compares it with the archaeological literature, then a good agreement can be found for *burial type*. Inhumation dominated for a long time, but was then replaced by cremation from the Middle Bronze Age onwards and massively in the Late Bronze Age. The calculated proportions for *burial construction* could easily be dismissed as completely worthless due to the negligibly small amount of data, but in fact the proxy indicates the right tendency for early and middle Bronze Age: The dominance of flat graves was replaced by mounds. The data set does not allow any statements about the Late Bronze Age.

`r a("Northeastern France")`:
*Northeastern France* comprises the French administrative regions Grand-Est and Bourgogne-Franche-Comté, the east of the German Baden-Württemberg, Rhineland-Palatinate and Saarland, as well as Luxembourg and Wallonia in Belgium. Accumulations of ^14^C data are only available from Württemberg and towards the Paris Basin. The amount of data is low overall, but very low in the early and early Middle Bronze Age. For *burial type*, after all, limited statements are possible after this time window, *burial construction* on the other hand must remain obscure. The prediction of the relative development of these variables is correspondingly error-prone. Still the parallel existence of cremation and body burial in early and middle Bronze Age is correctly reflected. However, this situation already existed in the Early Bronze Age and certainly developed more nuanced than the proxy might suggest. Contrary to the derivation from the ^14^C ratios, flat graves were quite common in Northeastern France.

`r a("Northern Germany")`:
The artificial region *Northern Germany* includes all federal states in East Germany, but also large parts of Lower Saxony and Schleswig-Holstein. The ^14^C data for this region come mainly from Saxony-Anhalt and neighbouring regions in Thuringia, Saxony and Lower Saxony. Also from the area north of Bremen some data are documented in Radon-B. The database does not contain as many dated graves from any other region as from northern Germany. Most of them date back to the Early Bronze Age, while the Middle Bronze Age from about 1800calBC to 1400calBC seems barren. Only for the Late Bronze Age there is a significant number of data documented again. The meta information on *burial type* and *burial construction* trace this development to the same extent. Since in Northern Germany influences from all four cardinal points come together and contribute significantly to the formation of local cultural phenomena, it is difficult to estimate the overall development of the primary variables here as well. The ^14^C proxy correctly represents *burial site* insofar as inhumations dominated in the Early Bronze Age and are replaced by cremations especially in the North from the Middle Bronze Age onwards. In the Late Bronze Age though, cremation was clearly more dominant than the data set suggests. With regard to *burial construction*, the current Radon-B stock records the basic dominance of flat graves well, but attributes more relevance to them in the Early and Middle Bronze Age than is usual in the archaeological assessment of the context.

`r a("Southern Scandinavia")`:
The region *Southern Scandinavia* includes all of Denmark with Jutland, Funen, Zeeland and all smaller islands as well as large parts of the east of southern Sweden, and coastal regions in Schleswig-Holstein and Mecklenburg-Western Pomerania. The ^14^C data are very well distributed across Jutland, Funen and the Swedish Skåne region. Since the Bronze Age in Scandinavia begins much later than in Central Europe and is measured by another chronological system, the diachronic developments cannot simply be compared by means of the tripartite structure of the Early, Middle and Late Bronze Ages. The period which is labeled Early Bronze Age due in Central Europe actually corresponds to the Scandinavian Late Neolithic. The majority of data from Scandinavia are from the period IB onwards. In this respect, the data density for almost the entire Bronze Age is good -- also with regard to the two variables *burial type* and *burial construction*. The Late Neolithic from 2200calBC is a desideratum. If one carries out the necessary mental shift -- the Central European Middle Bronze Age corresponds to the Nordic Early Bronze Age -- and then compares the proxy of the relative grave form development with the archaeological observations, then one finds a solid agreement: body graves were the rule in periods I and II (about 1800-1300calBC), but cremations also occurred sporadically. From period III onwards cremation burials gained the upper hand. In the Scandinavian Early Bronze Age burials in burial mounds were the absolute rule. In the Late Bronze Age, flat burial fields were added.

`r a("Benelux")`:
The Benelux region includes the Netherlands, northern Belgium and large parts of Lower Saxony and North Rhine-Westphalia. All these regions have ^14^C data on graves in Radon-B, but the data density in Belgium and the Netherlands is much higher. As in Southern Scandinavia, the Early Bronze Age is under-represented in the Benelux region as opposed to the Middle and Late Bronze Age, although in a slightly weaker form. While the information density of the *burial type* variable follows this course, *burial construction* is almost undocumented in the Late Bronze Age. As in the *Southern Germany* region, there is a serious difference between the information density of both variables. If one considers the relative development of the variables as they can be derived from the grave data in comparison to the archaeological state of research there is both deviation as well as agreement. Here, too, the study region includes different cultural phenomena with highly different burial customs: In the north and east of the Benelux region inhumations were the rule until the Late Bronze Age, while cremation dominated earlier south of the Meuse. The Radon-B proxy seems to overemphasise the dominance of cremation in the whole area. The relevance of hill burials in the Early and Middle Bronze Age, on the other hand, is correctly reflected, and the mixture between hill and flat graves in the context of urnfields, often resulting from the continued use of existing monumental sites, is also visible.

`r a("England")`:
The *England* region stretches across England up to York and Liverpool, incorporating parts of Wales and leaving out Cornwall and Devon in the southwest. It also touches a narrow coastal strip on the mainland, from which no ^14^C data are documented. The available data are concentrated in the southeast of Britain, but are well distributed there. The diachronic data availability is relatively good in the early and middle Bronze Age, but then decreases in the late Bronze Age. *burial type* and *burial construction* behave accordingly. The relative development derived from the ^14^C data indicates the correct trend for *burial type*: Cremation graves became more and more important during the Bronze Age, while inhumations were practiced for a long time in parallel. However, the data set emphasizes the dominance of the cremation graves too much. With regard to the variables *burial construction*, the basic dominance of burial mounds corresponds perfectly to the archaeological narrative, but nevertheless -- especially in the Late Bronze Age -- flat grave fields also occurred, which the data set wrongly excludes.

>>>>

General observations: **TODO**

`r a("Southeastern Central Europe")`^[Amount of graves in Radon-B for this region.]:
The artificial study region *Southeastern Central Europe* comprises the eastern part of Austria (Upper Austria, Lower Austria, Styria, Burgenland), the northwest of Hungary, Western Slovakia, Moravia and large parts of Bohemia (Figure \ref{map} B). The density of ^14^C data on graves in Radon-B is low -- individual dates are available from the various regions, but large areas are without evidence. Diachronically the amount of data is consistently low and reaches a minimum in the Middle Bronze Age. While for the *burial type* variable a constant flow of information is given, the number of graves without information on *burial construction* increases in the Late Bronze Age to such an extent that no meaningful overview can be derived. The comparison with archaeological observations is complicated by the fact that the region intersects diverse geographical and cultural areas. In the southeast of Central Europe, cremation and bodily burial already existed side by side in the Early Bronze Age, and the local tradition remained stable in this respect also due to the expansion of the burial mound tradition in the Middle Bronze Age -- with a slight overall increase in cremation graves. The Urnfield period brought the transition to the absolute dominance of cremation burial. The development of the *burial type* variable in the Radon-B proxy shows this development correctly. The information from ^14^C data on *burial construction* correctly traces the relevance of flat burials in the Early Bronze Age, but the latter are then overemphasized in the Middle Bronze Age compared to the construction of burial mounds. In the Late Bronze Age, burials on flat burial grounds were frequent, although burial mounds of the Tumulus Culture were also further documented. This impact cannot be correctly deduced from Radon-B -- certainly due to the small amount of data [@lubosCzechLandsAustria2013; @markovaSlovakiaHungary2013]. \newline

`r a("Poland")`:
The region *Poland* completely covers the centre and western part of today's Poland. However, the ^14^C data from this region are almost exclusively from the voivodships of Kajuwien-Pommern, Lodsch and Lower Silesia. The amount of data is small overall, with significantly more data from the Early Bronze Age up to about 1700calBC. Almost no information is available from the Middle and Late Bronze Ages. This applies to both the *burial type* and the *burial construction* variables, the former being slightly better documented. The relative development of *burial type* approximates the *real* conditions -- i.e. according to the classical, archaeological narrative: inhumations were actually significantly more frequent in western Poland until the Late Bronze Age, while cremations did not occur until the Urnfield Period. Also before, cremation graves had been created sometimes, but in general the tendency recorded by Radon-B is correct. In contrast, *burial construction* is misrepresented: burial mounds also played an important role in the Middle Bronze Age [@czebreszukBronzeAgePolish2013; @dabrowskiAeltereBronzezeitPolen2004]. \newline

`r a("Southern Germany")`:
The artificial district region *Southern Germany* covers Bavaria completely, but also includes most of Baden-Württemberg and Thuringia as well as extensive areas in Moravia and Western Austria. No ^14^C data are available from the latter regions. The majority of the available data comes from the Upper Palatinate, Lower Bavaria and Swabia as well as (in Württemberg) the administrative district of Tübingen. For the Early Bronze Age a remarkable amount of data from Southern Germany are preserved in Radon-B, but this amount decreases in the course of the Middle Bronze Age and leaves the Late Bronze Age almost without dated graves. The *burial type* variable is quite well understood -- although disproportionately much more of the early body graves were recorded than urnfield burials. This wealth of information contrasts sharply with the situation regarding the *burial construction* variable, for which almost no information from this area is available throughout the Bronze Age. If one considers the relative development calculated from Radon-B and compares it with the archaeological literature, then a good agreement can be found for *burial type*. Inhumation dominated for a long time, but was then replaced by cremation from the Middle Bronze Age onwards and massively in the Late Bronze Age. The calculated proportions for *burial construction* could easily be dismissed as completely worthless due to the negligibly small amount of data, but in fact the proxy indicates the right tendency for early and middle Bronze Age: The dominance of flat graves was replaced by mounds. The data set does not allow any statements about the Late Bronze Age [@falkensteinWandelBestattungssittenHugelgraber2017; @jockenhovelGermanyBronzeAge2013]. \newline

`r a("Northeastern France")`:
*Northeastern France* comprises the French administrative regions Grand-Est and Bourgogne-Franche-Comté, the east of the German Baden-Württemberg, Rhineland-Palatinate and Saarland, as well as Luxembourg and Wallonia in Belgium. Accumulations of ^14^C data are only available from Württemberg and towards the Paris Basin. The amount of data is low overall, but very low in the early and early Middle Bronze Age. For *burial type*, after all, limited statements are possible after this time window, *burial construction* on the other hand must remain obscure. The prediction of the relative development of these variables is correspondingly error-prone. Still the parallel existence of cremation and body burial in early and middle Bronze Age is correctly reflected. However, this situation already existed in the Early Bronze Age and certainly developed more nuanced than the proxy might suggest. Contrary to the derivation from the ^14^C ratios, flat graves were quite common in Northeastern France [@mordantBronzeAgeFrance2013]. \newline

`r a("Northern Germany")`:
The artificial region *Northern Germany* includes all federal states in East Germany, but also large parts of Lower Saxony and Schleswig-Holstein. The ^14^C data for this region come mainly from Saxony-Anhalt and neighbouring regions in Thuringia, Saxony and Lower Saxony. Also from the area north of Bremen some data are documented in Radon-B. The database does not contain as many dated graves from any other region as from northern Germany. Most of them date back to the Early Bronze Age, while the Middle Bronze Age from about 1800calBC to 1400calBC seems barren. Only for the Late Bronze Age there is a significant number of data documented again. The meta information on *burial type* and *burial construction* trace this development to the same extent. Since in Northern Germany influences from all four cardinal points come together and contribute significantly to the formation of local cultural phenomena, it is difficult to estimate the overall development of the primary variables here as well. The ^14^C proxy correctly represents *burial site* insofar as inhumations dominated in the Early Bronze Age and are replaced by cremations especially in the North from the Middle Bronze Age onwards. In the Late Bronze Age though, cremation was clearly more dominant than the data set suggests. With regard to *burial construction*, the current Radon-B stock records the basic dominance of flat graves well, but attributes more relevance to them in the Early and Middle Bronze Age than is usual in the archaeological assessment of the context [@hofmannRituelleUmgangMit2008; @jockenhovelGermanyBronzeAge2013]. \newline

`r a("Southern Scandinavia")`:
The region *Southern Scandinavia* includes all of Denmark with Jutland, Funen, Zeeland and all smaller islands as well as large parts of the east of southern Sweden, and coastal regions in Schleswig-Holstein and Mecklenburg-Western Pomerania. The ^14^C data are very well distributed across Jutland, Funen and the Swedish Skåne region. Since the Bronze Age in Scandinavia begins much later than in Central Europe and is measured by another chronological system, the diachronic developments cannot simply be compared by means of the tripartite structure of the Early, Middle and Late Bronze Ages. The period which is labeled Early Bronze Age due in Central Europe actually corresponds to the Scandinavian Late Neolithic. The majority of data from Scandinavia are from the period IB onwards. In this respect, the data density for almost the entire Bronze Age is good -- also with regard to the two variables *burial type* and *burial construction*. The Late Neolithic from 2200calBC is a desideratum. If one carries out the necessary mental shift -- the Central European Middle Bronze Age corresponds to the Nordic Early Bronze Age -- and then compares the proxy of the relative grave form development with the archaeological observations, then one finds a solid agreement: body graves were the rule in periods I and II (about 1800-1300calBC), but cremations also occurred sporadically. From period III onwards cremation burials gained the upper hand. In the Scandinavian Early Bronze Age burials in burial mounds were the absolute rule. In the Late Bronze Age, flat burial fields were added [@thraneScandinavia2013]. \newline

`r a("Benelux")`:
The Benelux region includes the Netherlands, northern Belgium and large parts of Lower Saxony and North Rhine-Westphalia. All these regions have ^14^C data on graves in Radon-B, but the data density in Belgium and the Netherlands is much higher. As in Southern Scandinavia, the Early Bronze Age is under-represented in the Benelux region as opposed to the Middle and Late Bronze Age, although in a slightly weaker form. While the information density of the *burial type* variable follows this course, *burial construction* is almost undocumented in the Late Bronze Age. As in the *Southern Germany* region, there is a serious difference between the information density of both variables. If one considers the relative development of the variables as they can be derived from the grave data in comparison to the archaeological state of research there is both deviation as well as agreement. Here, too, the study region includes different cultural phenomena with highly different burial customs: In the north and east of the Benelux region inhumations were the rule until the Late Bronze Age, while cremation dominated earlier south of the Meuse. The Radon-B proxy seems to overemphasise the dominance of cremation in the whole area. The relevance of hill burials in the Early and Middle Bronze Age, on the other hand, is correctly reflected, and the mixture between hill and flat graves in the context of urnfields, often resulting from the continued use of existing monumental sites, is also visible [@arnoldussenBronzeAgeSettlements2008; @drenthMoundsDeadFuneray2005; @lohofTraditionChangeBurial1994]. \newline

`r a("England")`:
The *England* region stretches across England up to York and Liverpool, incorporating parts of Wales and leaving out Cornwall and Devon in the southwest. It also touches a narrow coastal strip on the mainland, from which no ^14^C data are documented. The available data are concentrated in the southeast of Britain, but are well distributed there. The diachronic data availability is relatively good in the early and middle Bronze Age, but then decreases in the late Bronze Age. *burial type* and *burial construction* behave accordingly. The relative development derived from the ^14^C data indicates the correct trend for *burial type*: Cremation graves became more and more important during the Bronze Age, while inhumations were practiced for a long time in parallel. However, the data set emphasizes the dominance of the cremation graves too much. With regard to the variables *burial construction*, the basic dominance of burial mounds corresponds perfectly to the archaeological narrative, but nevertheless -- especially in the Late Bronze Age -- flat grave fields also occurred, which the data set wrongly excludes [@robertsBritainIrelandBronze2013]. \newline
